---
title: "DS311  EDA"
author: "Gabrielle Salamanca"
date: "April 23, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Dataset

First, let's call the necessary libraries:

```{r}
library(readr)
library(tibble)
```

Now, let's load the data and check if it loaded correctly:

```{r}
# data
ames <- read_csv("data/ames.csv", show_col_types = FALSE, col_types = cols())

# checking
if (!is.data.frame(ames)) {
  stop("ames is not a dataframe.")
}

if (nrow(ames) != 1460) {
  stop("ames does not have the correct number of rows.")
}

if (ncol(ames) != 81) {
  stop("ames does not have the correct number of columns.")
}
```

Let's inspect the contents of the dataframe

```{r}
# inspect
ames
str(ames)
```

## Explore Data Distributions

```{r}
ames.hist <- function(ames, column, title, xlabel, ylabel) 
{
  data <- ames[[column]]
  mean <- mean(data)
  hist(data, breaks = "FD", main = title, xlab = xlabel, ylab = ylabel, col = "darkmagenta")
  abline(v = mean, col = "goldenrod1")
}
# call function
ames.hist(ames, "SalePrice", "Distribution of Sale Prices", "Sale Price", "Number of Houses")
```

```{r}
mean <- mean(ames$SalePrice)
med <-  median(ames$SalePrice)
stdev <- sd(ames$SalePrice)
# print
cat("Mean: $", mean, "\n")
cat("Median: $", med, "\n")
cat("Standard Dev: $", stdev)
```

We see that the distribution looks like a log normal. It seems most of the houses cluster around 163000 USD. However, the higher-end homes are pulling up the average to 180921.2 USD.

## Explore Differences between Subsets

## Explore Correlations

## Engineer and Explore a New Feature
