---
title: "DS311  EDA"
author: "Gabrielle Salamanca"
date: "April 23, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the Dataset

First, let's call the necessary libraries:

```{r}
library(readr)
library(tibble)
library(caret)
library(dplyr)
```

Now, let's load the data and check if it loaded correctly:

```{r}
# data
ames <- read_csv("data/ames.csv", show_col_types = FALSE, col_types = cols())

# checking
if (!is.data.frame(ames)) {
  stop("ames is not a dataframe.")
}

if (nrow(ames) != 1460) {
  stop("ames does not have the correct number of rows.")
}

if (ncol(ames) != 81) {
  stop("ames does not have the correct number of columns.")
}
```

Let's inspect the contents of the dataframe

```{r}
# inspect
ames
str(ames)
```

## Explore Data Distributions

### 1. Sale Price

```{r}
ames.hist <- function(ames, column, title, xlabel, ylabel) 
{
  data <- ames[[column]]
  mean <- mean(data)
  hist(data, breaks = "FD", main = title, xlab = xlabel, ylab = ylabel, col = "darkmagenta")
  abline(v = mean, col = "goldenrod1")
}
# call function
ames.hist(ames, "SalePrice", "Distribution of Sale Prices", "Sale Price", "Number of Houses")
```

```{r}
mean <- mean(ames$SalePrice)
med <-  median(ames$SalePrice)
stdev <- sd(ames$SalePrice)
# print
cat("Mean: $", mean, "\n")
cat("Median: $", med, "\n")
cat("Standard Dev: $", stdev)
```

Looks like a log normal distribution. Most houses in this sample are clustered around the median value of \$163,000, but the higher-end homes are pulling the mean up to over \$180,000

### 2. Total Rooms Above Grade

```{r}
ames.hist(ames, "TotRmsAbvGrd", 
          "Distribution of Total Rooms Above Grade", 
          "Total Rooms (Does Not Include Bathrooms)",
          "Number of Houses")
```

```{r}
mean <- mean(ames$TotRmsAbvGrd)
med <-  median(ames$TotRmsAbvGrd)
stdev <- sd(ames$TotRmsAbvGrd)
# print
cat("Mean:", mean, "\n")
cat("Median:", med, "\n")
cat("Standard Dev:", stdev)
```

The number of rooms in houses is approximately normally distributed, with a mean and median around 6 rooms. There are some houses with twice as many rooms as the average, but overall the distribution is less skewed than the sale price distribution

### 3. Overall Condition

```{r}
ames.hist(ames, "OverallCond",
          "Distribution of Overall Condition of Houses on a 1-10 Scale",
          "Condition of House", "Number of Houses")
# vars
mean <- mean(ames$OverallCond)
med <-  median(ames$OverallCond)
stdev <- sd(ames$OverallCond)
```

```{r}
# print
cat("Mean:", mean, "\n")
cat("Median:", med, "\n")
cat("Standard Dev:", stdev)
```

Most homes have a condition of 5. It seems like we should treat this as a categorical rather than numeric variable, since the difference between conditions is so abrupt

## Explore Differences between Subsets

```{r}
below <- ames[ames$OverallCond < 5, ]
avg <- ames[ames$OverallCond == 5, ]
above <- ames[ames$OverallCond > 5, ]

# plot setup
par(mfrow = c(1,1))
plot.new()
plot.window(xlim = c(min(ames$SalePrice), max(ames$SalePrice)), ylim = c(0, max(hist(ames$SalePrice, breaks = bins)$counts)))
# bins
bins <- seq(min(ames$SalePrice), max(ames$SalePrice), length.out = 21)

# plot
hist(
  above$SalePrice,
  breaks = bins,
  col = rgb(0,1,1,alpha=0.5),
  xlab = "Sale Price",
  ylab = "Number of Houses",
  freq = FALSE,
  main = "Distributions of Sale Price Grouped by Condition"
)
hist(
  avg$SalePrice,
  breaks = bins,
  col = rgb(0.5,0.5,0.5,alpha=0.3),
  add = TRUE,
  freq = FALSE
)
hist(
  below$SalePrice,
  breaks = bins,
  col = rgb(1,1,0,alpha=0.5),
  add = TRUE,
  freq = FALSE
)

# legend
legend(
  "topright",
  legend = c("above average condition", "average condition", "below average condition"),
  fill = c(rgb(0, 1, 1,alpha = 0.5), 
           rgb(0.5, 0.5, 0.5, alpha = 0.3), 
           rgb(1, 1, 0, alpha = 0.5))
)
```

First, we note again that the majority of the houses have average condition, then about 1/3 have above average condition, then less than 10% have below average condition.

As we might expect, the average condition therefore contains houses across a broader spectrum of the sale price range than either the below-average or above-average houses.

Another unsurprising finding is that below-average condition houses have a price distribution that is much lower than average or above-average condition houses.

But what might be surprising is that above-average condition houses do not seem to have higher average sale prices than average condition houses. In fact, above-average condition houses seem more clustered around a particular price range, especially the \$100,000 to \$200,000 range, whereas average condition houses are more frequent above \$200,000. We might want to investigate further to understand what kinds of houses are rated as above-average condition, since this goes against a standard assumption that better condition would mean higher cost.

## Explore Correlations

### 1. Most positively correlated

```{r}
num <- df %>%
select_if(is.numeric) %>%
select(-SalePrice)

correlations <- c()
for (col in colnames(num)) {
corr <- cor(ames[[col]], ames[["SalePrice"]])
correlations <- c(correlations, corr)
}
max_corr_value <- max(correlations)
max_corr_column <- colnames(num)[which.max(correlations)]
cat("Most Positively Correlated Column:", max_corr_column, "\n")
cat("Maximum Correlation Value:", max_corr_value, "\n")
```


## Engineer and Explore a New Feature
